---
title: "College Majors, Gender Composition, and Early-Career Earnings"
author: "Peace Njoku"
format: html
execute:
  echo: false
---

```{r}
#| message: false
library(tidyverse)
library(gtsummary)

# read in the cleaned data saved by clean-data.R
majors_cats <- read_rds(here::here("data", "clean", "majors_clean.rds"))
```

Description of the Data

The data comes from FiveThirtyEight's *The Economic Guide To Picking A College
Major*, built from the American Community Survey 2010--2012 Public Use Microdata
Sample. The file is available at
<https://github.com/fivethirtyeight/data/tree/master/college-majors>.

Each row is one undergraduate major, all findings below describe majors.
The dataset covers `r nrow(majors_cats)` majors across `r n_distinct(majors_cats$Major_category)`
broad Major categories. 
Majors were split at a 50% threshold into those where women
make up under half of graduates and those where women make up half or more.

Majors where women make up at least half of graduates report substantially lower
median earnings (@table1).

```{r}
#| label: table1
#| tbl-cap: "TABLE 1: Characteristics of undergraduate majors, by gender"
#| message: false


table1 <- tbl_summary(
  majors_cats,
  by = women_cat,
  include = c(Median, unempl_pct, Total),
  label = list(
    Median ~ "Median earnings (USD)",
    unempl_pct ~ "Unemployment rate (%)",
    Total ~ "Number of graduates"
  ),
  missing_text = "Missing")|>
  
  # test whether the two groups differ
  add_p(test = all_continuous() ~ "t.test")|>
  bold_labels() |>
  
  # rename and bold the first and last column headers
  modify_header(label = "**Characteristic**", p.value = "**P-value**")

table1
  
```

Among women dominated majors, the median earning was
`r inline_text(table1, variable = "Median", column = "stat_2", pattern = "{median}")` USD.


