---
title: "College Majors, Gender Composition, and Early-Career Earnings"
author: "Peace Njoku"
format: html
execute:
  echo: false
---

```{r}
#| message: false
library(tidyverse)
library(gtsummary)

# read in the cleaned data saved by clean-data.R
majors_cats <- read_rds(here::here("data", "clean", "majors_clean.rds"))
```
##Introduction

Some college majors are studied mostly by women, others mostly by men. This analysis asks a simple question: Do majors with more women in them lead to lower pay?

##Description of the Data

The data comes from FiveThirtyEight's *The Economic Guide To Picking A College
Major*, built from the American Community Survey 2010--2012 Public Use Microdata
Sample. The file is available at
<https://github.com/fivethirtyeight/data/tree/master/college-majors>.

Each row is one undergraduate major not a person. Each row tells you what a typical graduate of that major earns, not what any particular graduate earns.
The dataset covers `r nrow(majors_cats)` majors across `r n_distinct(majors_cats$Major_category)`
broad fields like engineering, health and education. 

##Data Analysis

Before comparing groups, here is what these majors look like overall (@tbl-zero). Each number describes the median of all of them. 

```{r}
#| label: tbl-zero
#| tbl-cap: "Characteristics of undergraduate majors"
#| message: false

table0 <- tbl_summary(
  majors_cats,
  include = c(women_pct, Median, unempl_pct),
  label = list(
    women_pct ~ "Women Graduates(%)",
    Median ~ "Median earnings (USD)",
    unempl_pct ~ "Unemployment rate (%)"
  ),
  statistic = all_continuous() ~ "{median}",
  missing_text = "Missing"
) |>
  bold_labels()

table0
```

So the typical major is `r inline_text(table0, variable = "women_pct", column = "stat_0")` percent women dominated, pays about $`r inline_text(table0, variable = "Median", column = "stat_0", pattern = "{median}")` USD a year, and has about `r inline_text(table0, variable = "unempl_pct", column = "stat_0")` percent of its graduate out of work.  

Majors were sorted into two groups based on the share of graduates who are women. Where women make up less than half the major is counted as 'Male Dominated', where they make up half or more, it is counted as 'Female-dominated'.

Looking at it this way, we see the two groups earn very different amounts, majors where women make up at least half of graduates report substantially lower
median earnings (@tbl-one).

```{r}
#| label: tbl-one
#| tbl-cap: "Characteristics of undergraduate majors, by gender"
#| message: false


table1 <- tbl_summary(
  majors_cats,
  by = women_cat,
  include = c(Median, unempl_pct, Total),
  label = list(
    Median ~ "Median earnings (USD)",
    unempl_pct ~ "Unemployment rate (%)",
    Total ~ "Number of graduates"
  ),
  missing_text = "Missing")|>
  
  # test whether the two groups differ
  add_p(test = all_continuous() ~ "t.test")|>
  bold_labels() |>
  
  # rename and bold the first and last column headers
  modify_header(label = "**Characteristic**", p.value = "**P-value**")

table1
  
```

Among women dominated majors, the typical graduate earns about $`r inline_text(table1, variable = "Median", column = "stat_2", pattern = "{median}")` USD.

In majors where men dominate, it's about $`r inline_text(table1, variable = "Median" , column = "stat_1", pattern = "{median}")` USD a substantial difference.

The unemployment row is almost identical between the two groups. Graduates of women-dominated majors find jobs just as easily, they simply get paid less for them. 

The regression uses the actual percentage and asks how much earnings change as the share of women goes up.(@tbl-two).

```{r}
#| label: tbl-two
#| tbl-cap: "Linear regression of median earnings on gender composition"

#linear model 
linear_model <- lm(Median ~ women_pct + unempl_pct,
                   data = majors_cats)

#regression table
table2 <- tbl_regression(
  linear_model,
  intercept = TRUE,
  label = list(
    women_pct ~ "Women (%)",
    unempl_pct ~ "Unemployment rate (%)"
               )
) |>
  bold_labels()

table2
```

For every one percent of women in a major, the typical pay changes by $`r inline_text(table2, variable = "women_pct", pattern = "{estimate}")` USD.

The model estimates that for every one percentage point more women in a major,
typical pay is about $304 lower. Put another way, a major with 10 percent more
women is expected to pay roughly $3,000 less. The p-value is below 0.001,
which means a pattern this strong would be very unlikely to appear by chance
alone. The unemployment rate, by contrast, has a p-value of 0.2 and a range
that includes zero, so this data cannot say whether it is related to pay at all.

Each dot in (@fig-one) is one major, the pattern as you move right towards majors with more women, the dots drift downward.

```{r}
#| label: fig-one
#| fig-cap: "Median earnings and gender composition, by major category"
#| warning: false

earnings_plot <- ggplot(majors_cats,
                        aes(x = women_pct, y = Median,
                            color = Major_category)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "black") +
  scale_y_continuous(labels = scales::comma) + #to correct the scientific notion
  labs(subtitle = "Each point is one undergraduate major",
       caption = "Source: FiveThirtyEight, American Community Survey 2010-2012",
      x = "Graduates who are women (%)",
       y = "Median earnings (USD)",
       color = "Major category"
  ) +
  theme_minimal()

earnings_plot

```

##Conclusion

There is a clear pattern here, but a pattern isn't a cause. This analysis cannot tell us why majors with more women pay less, it could be various reasons which the data cannot separate. 

It is also important that pay here covers *everyone* who graduated from a major,
men and women together. The data does not report what men and women earn
separately. So this is not a finding about women being paid less than men for
the same work. It is a finding about which fields pay well and who ends up in
them.

Finally, each row is a major, not a person, so nothing here describes what any
individual graduate earns.
